---
title: "Find"
subtitle: "Search every function in the animovement packages, the formats they read, and the methods they offer."
toc: false
---

```{r}
#| label: setup
#| include: false
library(animovement)

pkgs <- c(
  "anicore", "aniread", "anicheck", "aniprocess",
  "anispace", "animetric", "anivis", "animovement"
)

reference_url <- function(package, topic) {
  sprintf("https://animovement.dev/%s/reference/%s.html", package, topic)
}

link <- function(text, url) sprintf('<a href="%s">%s</a>', url, text)

# The parts of an Rd file with a given tag, as plain text
rd_field <- function(rd, tag) {
  tags <- vapply(rd, function(x) attr(x, "Rd_tag"), "")
  unlist(lapply(rd[tags == tag], function(x) paste(unlist(x), collapse = "")))
}

# One row per help topic that documents exported functions
topics <- lapply(pkgs, function(package) {
  db <- tools::Rd_db(package)
  exports <- getNamespaceExports(package)
  rows <- lapply(names(db), function(file) {
    rd <- db[[file]]
    if ("internal" %in% rd_field(rd, "\\keyword")) {
      return(NULL)
    }
    functions <- intersect(rd_field(rd, "\\alias"), exports)
    if (length(functions) == 0) {
      return(NULL)
    }
    description <- paste(rd_field(rd, "\\description"), collapse = " ")
    stage <- regmatches(
      description,
      regexpr("#(experimental|superseded|deprecated)", description)
    )
    data.frame(
      package = package,
      topic = sub("[.]Rd$", "", file),
      functions = paste(functions, collapse = ", "),
      title = trimws(gsub("\\s+", " ", rd_field(rd, "\\title"))),
      stage = if (length(stage)) sub("#", "", stage) else ""
    )
  })
  do.call(rbind, rows)
})
topics <- do.call(rbind, topics)

search_table <- function(data, page_length = 25) {
  DT::datatable(
    data,
    rownames = FALSE,
    escape = FALSE,
    style = "bootstrap5",
    class = "table table-sm",
    options = list(pageLength = page_length, autoWidth = FALSE)
  )
}
```

Looking for a function, but not sure which package it's in? Every function the
*animovement* packages export is below, with a link to its help page. Type in
the search box to filter by name, package or description.

::: {.panel-tabset}
## All functions

```{r}
#| label: all-functions
#| echo: false
topics |>
  dplyr::transmute(
    Function = link(functions, reference_url(package, topic)),
    Package = link(package, sprintf("https://animovement.dev/%s/", package)),
    Description = title,
    Stage = stage
  ) |>
  search_table()
```

Functions marked *experimental* may still change. See the
[lifecycle stages](https://lifecycle.r-lib.org/articles/stages.html) for what
each stage means.

## Data sources

`read_dataset()` reads all of these formats, working out the source from the
file itself. Each source also has its own reader, whose help page documents
the options for that format.

```{r}
#| label: sources
#| echo: false
get_supported_sources() |>
  dplyr::transmute(
    Source = source,
    Reader = link(reader, reference_url("aniread", reader)),
    `File types` = vapply(suffix, paste, "", collapse = ", ")
  ) |>
  search_table()
```

## Processing methods

The `*_across()` functions apply a method to every position column of an
anipoint, and the `*_with()` functions do the same for a single column. Each
method is also a function of its own.

```{r}
#| label: methods
#| echo: false
methods_of <- function(fun) eval(formals(fun)$method)

methods <- rbind(
  data.frame(
    Step = "Mask",
    Method = methods_of(aniprocess::mask_na_across),
    prefix = "mask_na_"
  ),
  data.frame(
    Step = "Fill",
    Method = methods_of(aniprocess::replace_na_across),
    prefix = "replace_na_"
  ),
  data.frame(
    Step = "Smooth",
    Method = methods_of(aniprocess::filter_across),
    prefix = "filter_"
  )
)
methods$fun <- paste0(methods$prefix, methods$Method)
methods <- merge(
  methods,
  topics[topics$package == "aniprocess", c("functions", "topic", "title")],
  by.x = "fun",
  by.y = "functions",
  all.x = TRUE,
  sort = FALSE
)
methods <- methods[order(match(methods$Step, c("Mask", "Fill", "Smooth"))), ]

methods |>
  dplyr::transmute(
    Step,
    Method = sprintf("<code>\"%s\"</code>", Method),
    Function = link(fun, reference_url("aniprocess", topic)),
    Description = title
  ) |>
  search_table()
```
:::
