---
title: "Calculate metrics"
description: "Turn clean positions into speed, acceleration, heading and distance travelled — then summarise them."
order: 5
---

```{r}
#| label: setup
#| include: false
library(animovement)
knitr::opts_chunk$set(
  comment = "#>",
  fig.width = 7, fig.height = 4.5, fig.align = "center",
  dpi = 150, out.width = "100%"
)
```

Positions are rarely the answer to a research question. Speed, distance
travelled, how much an animal turned and when — those are.
[animetric](https://animovement.dev/animetric) derives them from a clean
aniframe.

```{r}
#| label: pipeline
#| warning: false
library(animovement)

clean <- read_sleap(get_sample_data("sleap")) |>
  set_sampling_rate(30) |>
  filter_na_across("roi", x_min = 300, x_max = 1010, y_min = 0, y_max = 690) |>
  filter_na_across("confidence", threshold = 0.6) |>
  replace_na_across("linear", max_gap = 5) |>
  filter_across("sgolay", window_width = 11, order = 3)
```

## Kinematics {#sec-kinematics}

`calculate_kinematics()` takes the aniframe and adds a column for each
derived quantity. No arguments to configure — it reads the coordinate system,
the sampling rate and the grouping from the metadata:

```{r}
#| label: kinematics
kinematics <- calculate_kinematics(clean)

setdiff(names(kinematics), names(clean))
```

Thirteen new columns, in two families. Translational: `speed`, `acceleration`,
`path_length`, and the per-axis components `v_x`, `v_y`, `a_x`, `a_y`.
Rotational: `heading` and its unwrapped twin, `angular_speed`,
`angular_velocity`, `angular_acceleration`, `angular_path_length`. For 3-D data
you get the translational family only.

```{r}
#| label: kin-print
kinematics
```

::: {.callout-important}
## Check your units before you believe a number
Speed here is **pixels per second** — pixels because that's what the video gave
us, per second because we set the sampling rate when reading. Get the sampling
rate wrong and every one of these numbers is wrong by the same factor. If you've
calibrated the recording, `set_unit_space()` before this step and you'll get
real-world units out.
:::

## Look at it {#sec-plot}

`plot_timeseries()` draws any column against the index:

```{r}
#| label: speed-plot
#| fig-height: 3.5
#| fig-cap: "Speed of the mouse's centre over ten minutes, in pixels per second."
kinematics |>
  dplyr::filter(keypoint == "centre") |>
  plot_timeseries(variable = "speed")
```

Bursts of movement separated by long still stretches — the shape you'd expect
from a mouse on a plus maze. Note the spike in the first seconds: the recording
starts before the tracking settles, and it's the kind of artefact worth going
back and masking with `filter_na_speed()` rather than explaining away.

## Summarise {#sec-summarise}

Per-frame kinematics are the raw material; usually you want one row per
individual. `summarise_kinematics()` collapses each group, defaulting to median
and MAD because movement data is rarely symmetric enough for mean and SD to
describe it well:

```{r}
#| label: summarise
summarise_kinematics(kinematics)
```

One row per keypoint here, since that's how this recording is grouped. The snout
moves fastest — it swings as the mouse turns, so it covers more ground than the
body it's attached to. Pass `measures = "mean_sd"` if you'd rather have mean and
standard deviation.

## Beyond kinematics {#sec-more}

The same aniframe feeds the rest of animetric:

- `calculate_tortuosity()` — how convoluted a path is, over sliding windows;
  `compute_straightness()`, `compute_sinuosity()` and `compute_emax()` give the
  individual indices.
- `calculate_nnd()` — distance to the *n*-th nearest neighbour, for recordings
  with several individuals.
- `add_centroid()` — add a derived centroid keypoint, e.g. the middle of a set
  of body points.
- `mean_angle()` and `median_angle()` — circular statistics, for angles where an
  ordinary mean would put the average of 350° and 10° at 180°.

And when the question is about space rather than time,
[anispace](https://animovement.dev/anispace) transforms the coordinates
themselves — `transform_to_egocentric()` to re-express the world from the
animal's own point of view, `map_to_polar()` and friends to change coordinate
system, `rotate_coords()` and `translate_coords()` to align recordings.

## Where next {#sec-next}

You've been through the whole arc: read, check, process, measure. From here:

- **[Packages](../packages/index.qmd)** — each package's own site, with the full
  function reference and examples.
- **[Learn](../learn/index.qmd)** — worked examples and end-to-end tutorials.
- **[Help](../help/index.qmd)** — where to ask when something doesn't behave.
