---
title: "Calculate metrics"
description: "Turn clean positions into speed, direction of travel, body heading and distance travelled, then summarise them."
order: 5
---

```{r}
#| label: setup
#| include: false
library(animovement)
knitr::opts_chunk$set(
  comment = "#>",
  fig.width = 7, fig.height = 4.5, fig.align = "center",
  dpi = 150, out.width = "100%"
)
```

Positions are rarely the answer to a research question. Speed, distance
travelled, and how much an animal turned and when usually are.
[animetric](https://animovement.dev/animetric) derives them from a clean
anipoint. We start from the pipeline in the [previous guide](process-data.qmd),
with one addition: angles in degrees rather than radians, which are easier to
read.

```{r}
#| label: pipeline
#| warning: false
library(animovement)

clean <- read_dataset(get_sample_data("sleap")) |>
  set_metadata(sampling_rate = 30) |>
  convert_unit_time("s") |>
  convert_unit_angle("deg") |>
  mask_na_across("roi", x_min = 300, x_max = 1010, y_min = 0, y_max = 690) |>
  mask_na_across("confidence", threshold = 0.6) |>
  replace_na_across("linear", max_gap = 5) |>
  filter_across("sgolay", window_width = 11, order = 3)
```

## Kinematics {#sec-kinematics}

`calculate_kinematics()` adds a column for each derived quantity. There's
nothing to configure: it reads the coordinate system, the sampling rate, the
angle unit and the grouping from the metadata.

```{r}
#| label: kinematics
kinematics <- calculate_kinematics(clean)

setdiff(names(kinematics), names(clean))
```

The new columns come in two families:

- **How fast:** `speed`, `acceleration`, `path_length` (the distance travelled
  so far), and the velocity and acceleration along each axis, `v_x`, `v_y`,
  `a_x` and `a_y`.
- **Which way:** `course`, the direction of travel, and how it changes:
  `turning_rate`, `turning_speed`, `turning_acceleration`, and
  `cumulative_turning`. `course_unwrapped` is the same direction without the
  jump from 180° to −180°.

```{r}
#| label: kin-print
kinematics
```

::: {.callout-important}
## Check your units before you believe a number
Speed here is in **pixels per second**: pixels because that's what the video
gave us, and seconds because we set the sampling rate and converted the time.
Get the sampling rate wrong and every one of these numbers is wrong by the same
factor. If you've calibrated the recording, use `convert_unit_space()` before
this step to get real-world units.
:::

## Look at it {#sec-plot}

`plot_timeseries()` draws a column against time:

```{r}
#| label: speed-plot
#| fig-height: 3.5
#| fig-cap: "Speed of the mouse's centre over ten minutes, in pixels per second."
kinematics |>
  dplyr::filter(keypoint == "centre") |>
  plot_timeseries(variable = "speed")
```

Bursts of movement separated by long still stretches: the pattern you'd expect
from a mouse exploring a plus maze. A spike that looks too sudden to be real is
worth going back to mask, with the `"speed"` or `"hampel"` method of
`mask_na_across()`, rather than explaining away.

## Where it goes and where it faces {#sec-heading}

`course` is the direction a point is *moving*. That isn't always the direction
the animal is *facing*: a mouse can turn its body while barely moving, and
when it stops, there is no direction of travel at all.

Where the body points is its orientation, and `add_orientation()` works it out
from two points. Here, the line from the base of the tail to the snout gives
the mouse's heading:

```{r}
#| label: orientation
oriented <- clean |>
  add_orientation(from = "tail_base", to = "snout", level = "keypoint") |>
  calculate_kinematics()
```

The heading is attached to every keypoint, since it belongs to the body as a
whole. Over one second of a run, the two directions start apart while the
mouse turns, then come together as it runs straight:

```{r}
#| label: course-heading
#| message: false
oriented |>
  dplyr::filter(keypoint == "centre", time > 100, time < 101) |>
  dplyr::select(time, speed, course, heading)
```

## Summarise {#sec-summarise}

Per-frame kinematics are the raw material; usually you want one row per track.
animetric has two kinds of summary.

`summarise_aniframe()` summarises the distribution of each per-frame measure.
It defaults to the median and the median absolute deviation (MAD), because
movement data is rarely symmetric enough for a mean and standard deviation to
describe it well. Angles such as `course` and `heading` get circular
statistics, so the average of 350° and 10° comes out as 0°, not 180°.

```{r}
#| label: summarise
summarise_aniframe(oriented)
```

There's one row per keypoint, since that's how this recording is grouped. The
snout moves fastest: it swings as the mouse turns its head, so it covers more
ground than the body it's attached to. Pass `measures = "mean_sd"` if you'd
rather have the mean and standard deviation.

`summarise_path()` describes each track's path as a whole: the total distance,
the total turning, how far the animal ended up from where it started, and how
straight or winding the path was:

```{r}
#| label: summarise-path
summarise_path(clean)
```

## Beyond kinematics {#sec-more}

The same anipoint feeds the rest of animetric:

- `add_point()` adds a point derived from others, such as the centroid of the
  head from the snout and both ears.
- `calculate_tortuosity()` measures how winding a path is over sliding windows.
- `calculate_nnd()` finds the distance to the nearest neighbour, for
  recordings with several individuals.

When the question is about space rather than time,
[anispace](https://animovement.dev/anispace) transforms the coordinates
themselves. `transform_to_egocentric()` re-expresses positions from the
animal's own point of view, `map_to_polar()` and related functions change the
coordinate system, and `rotate_coords()` and `translate_coords()` align
recordings with each other.

## Where next {#sec-next}

You've been through the whole workflow: read, check, process, measure. From
here:

- **[Packages](../packages/index.qmd)** has each package's own site, with the
  full function reference and examples.
- **[Learn](../learn/index.qmd)** is for worked examples and end-to-end
  tutorials.
- **[Help](../help/index.qmd)** is where to ask when something doesn't behave.
