---
title: "Check your data"
description: "Find out how good your tracking really is (confidence, gaps, and whether the body keeps its shape) before changing anything."
order: 3
---

```{r}
#| label: setup
#| include: false
library(animovement)
knitr::opts_chunk$set(
  comment = "#>",
  fig.width = 7, fig.height = 4.5, fig.align = "center",
  dpi = 150, out.width = "100%"
)
```

Tracking data is never perfect. Keypoints get occluded, the tracker guesses,
and once a guess is written to file it looks exactly like a measurement: a pair
of plausible coordinates.

[anicheck](https://animovement.dev/anicheck) helps you tell them apart. Its
functions only *report*. Each takes an anipoint and returns a summary, leaving
your data untouched. Cleaning comes in the next guide, once you know what
needs cleaning.

We start from where the [previous guide](read-data.qmd) left off:

```{r}
#| label: read
library(animovement)

path <- get_sample_data("sleap")
movement <- read_dataset(path) |>
  set_metadata(sampling_rate = 30) |>
  convert_unit_time("s") |>
  set_structure(read_structure(path))
```

## How confident was the tracker? {#sec-confidence}

Pose-estimation tools score every point they place. `check_confidence()`
summarises those scores per keypoint:

```{r}
#| label: check-conf
check_confidence(movement)
```

The snout and the tail end are tracked far less confidently than the centre or
the ears. That makes sense, since they're the parts that disappear when the
mouse turns. To see the whole distribution rather than a summary, plot the
check:

```{r}
#| label: plot-conf
#| fig-cap: "Confidence per keypoint. The spike at zero is the tracker admitting it had no idea."
plot(check_confidence(movement))
```

This is the plot to choose a threshold from. Most of each keypoint's scores sit
near 1, with a separate clump at zero and a thin spread in between, so a cut
around 0.6 removes the failures without touching the good tracking. Where
exactly to cut is a judgement about your data, and this plot lets you make it
rather than guess.

::: {.callout-note}
## Confidence isn't a probability
Scores are whatever the tracker chose to emit. SLEAP's run past 1 in this
recording, and a 0.9 from one tool says nothing about a 0.9 from another. Read
the distribution, rather than borrowing a threshold from someone else's paper.
:::

## Where are the gaps? {#sec-gaps}

Missing values are the other half of the story. `check_na_gapsize()` reports how
many there are and, more usefully, whether they come as short dropouts or long
absences:

```{r}
#| label: check-gaps
check_na_gapsize(movement)
```

The overall percentage hides the shape of the problem. The snout alone has
hundreds of separate gaps, and the longest runs to nearly 200 frames. Plotting
one keypoint at a time shows the shape:

```{r}
#| label: plot-gaps
#| fig-height: 3.5
#| fig-cap: "Gap sizes for the snout. Most gaps are a handful of frames long."
movement |>
  dplyr::filter(keypoint == "snout") |>
  check_na_gapsize() |>
  plot()
```

The difference matters for the next step. A gap of one frame is worth
interpolating across, because the mouse can't have gone far. A gap of 200
frames is nearly seven seconds, and anything drawn across it is invention.

::: {.callout-tip}
## Plotting several groups at once
These check plots put each group in its own panel, so with six keypoints you'll
want a taller figure (`fig-height: 10` or so) before they become readable. One
keypoint at a time is usually easier to read anyway.
:::

## When do they happen? {#sec-timing}

Gaps clustered in one stretch mean something different from gaps spread evenly:
a lighting change, say, rather than a keypoint the model was never good at.
`check_na_timing()` splits the recording into intervals and shows how much of
each is missing:

```{r}
#| label: check-timing
#| fig-height: 3.5
#| fig-cap: "When the snout goes missing, interval by interval."
movement |>
  dplyr::filter(keypoint == "snout") |>
  check_na_timing() |>
  plot()
```

Here the gaps are spread across the whole recording, with a few worse patches.
That points to the mouse moving out of the camera's clear view, in the closed
arms, rather than to a broken stretch of recording. There's nothing to fix in
the setup, but it's worth remembering when you interpret behaviour in those
stretches.

## Does the body keep its shape? {#sec-segments}

A mouse doesn't stretch to twice its length between two frames, so a body
segment whose length jumps around points to a keypoint placed in the wrong
spot. With the skeleton attached, `check_segment_length()` compares each
segment's length in every frame with its usual length:

```{r}
#| label: check-segments
check_segment_length(movement)
```

```{r}
#| label: plot-segments
#| fig-height: 5
#| fig-cap: "Each segment's length relative to its usual length. The dashed lines mark 30% either way."
plot(check_segment_length(movement))
```

The body axis, from snout to centre to tail, holds its length well. The
segments from the snout to the ears don't: they're well off their usual length
in a large share of frames. The ears are small and close together, so they're
the easiest points for a tracker to misplace. Keep that in mind before using
the ears for anything precise, such as head direction.

## What we learned {#sec-summary}

What the checks mean for the [next guide](process-data.qmd):

- The maze occupies roughly x from 300 to 1010 and y from 0 to 690. Points
  outside it are impossible, so they can go.
- Confidence separates cleanly around **0.6**, so that's where we'll mask.
- Most gaps are a few frames long and safe to interpolate, but a handful run to
  nearly 200 frames and are not.
- The missing data is spread through the recording rather than concentrated in
  one broken stretch, so there's no reason to discard part of it.
- The ears are the least reliable part of the skeleton.

Next: [process the trajectories](process-data.qmd).
