---
title: "Check your data"
description: "Find out how good your tracking really is — confidence, gaps, and when they happen — before changing anything."
order: 3
---

```{r}
#| label: setup
#| include: false
library(animovement)
knitr::opts_chunk$set(
  comment = "#>",
  fig.width = 7, fig.height = 4.5, fig.align = "center",
  dpi = 150, out.width = "100%"
)
```

Tracking data is never perfect. Keypoints get occluded, the tracker guesses,
and a guess looks exactly like a measurement once it's written to file — a pair
of plausible coordinates, indistinguishable from the real thing.

[anicheck](https://animovement.dev/anicheck) exists to tell them apart. Its
functions only *report*: they take an aniframe and give you back a summary,
leaving your data untouched. Cleaning comes next, once you know what you're
cleaning.

```{r}
#| label: read
library(animovement)

movement <- read_sleap(get_sample_data("sleap")) |>
  set_sampling_rate(30)
```

## How confident was the tracker? {#sec-confidence}

Pose-estimation tools score every point they place. `check_confidence()`
summarises those scores per keypoint:

```{r}
#| label: check-conf
check_confidence(movement)
```

Already informative: the snout and tail end are tracked far less confidently
than the centre or the ears — which makes sense, they're the parts that
disappear when the mouse turns. Pair the check with `plot()` to see the whole
distribution rather than two numbers:

```{r}
#| label: plot-conf
#| fig-cap: "Confidence per keypoint. The spike at zero is the tracker admitting it had no idea."
plot(check_confidence(movement))
```

This is the plot that picks your threshold. The bulk of each keypoint sits near
1, with a separate clump at zero and a thin tail between — so a cut somewhere
around 0.6 removes the failures without touching the good tracking. Where
exactly to cut is a judgement call about your data, and this plot is how you
make it rather than guess it.

::: {.callout-note}
## Confidence isn't a probability
Scores are whatever the tracker chose to emit. SLEAP's run past 1 in this
recording, and a "0.9" from one tool means nothing about a "0.9" from another.
Read the distribution, don't import a threshold from someone else's paper.
:::

## Where are the gaps? {#sec-gaps}

Missing values are the other half of the story. `check_na_gapsize()` reports how
many there are and, more usefully, how they're distributed between short
dropouts and long absences:

```{r}
#| label: check-gaps
check_na_gapsize(movement)
```

Eight percent of this recording is missing, but that number hides the shape of
the problem: the snout alone accounts for 331 separate gaps, one of them 197
frames long. Plotting one keypoint at a time shows the shape clearly:

```{r}
#| label: plot-gaps
#| fig-height: 3.5
#| fig-cap: "Gap sizes for the snout. Most gaps are a handful of frames; the red bars count gaps, the blue bars count the missing values they add up to."
movement |>
  dplyr::filter(keypoint == "snout") |>
  check_na_gapsize() |>
  plot()
```

That distinction matters for the next step. A one-frame gap is worth
interpolating across — the mouse can't have gone far. A 197-frame gap is nearly
seven seconds, and anything you draw across it is invention.

::: {.callout-tip}
## Plotting several groups at once
These check plots put each group in its own panel, so with six keypoints you'll
want a taller figure — `fig-height: 10` or so — before they become readable.
One keypoint at a time is usually easier to read anyway.
:::

## When do they happen? {#sec-timing}

Gaps clustered in one stretch mean something different from gaps sprinkled
evenly — a lighting change, say, versus a keypoint the model was never good at.
`check_na_timing()` splits the recording into intervals and shows what fraction
of each is missing:

```{r}
#| label: check-timing
check_na_timing(movement)
```

```{r}
#| label: plot-timing
#| fig-height: 3.5
#| fig-cap: "When the snout goes missing. Nearly half of some early intervals, barely any of others."
movement |>
  dplyr::filter(keypoint == "snout") |>
  check_na_timing() |>
  plot()
```

Here it's spread across the whole recording, with a couple of bad patches near
the start and around eight minutes in — the mouse in the closed arms, out of
sight of the camera. Nothing to fix in the setup; something to be careful about
when interpreting behaviour from those stretches.

## What we learned {#sec-summary}

Three checks, three decisions for the [next guide](process-data.qmd):

- The maze occupies roughly x ∈ [300, 1010], y ∈ [0, 690]; the few points
  outside it are impossible, so they can go.
- Confidence separates cleanly around **0.6**, so that's where we'll mask.
- Most gaps are a few frames long and safe to interpolate; a handful run to
  nearly 200 frames and are not.
- The missing data is spread through the recording rather than concentrated in
  one broken stretch, so there's no reason to discard part of it.

Next: [process the trajectories](process-data.qmd).
