---
title: "Check your data"
description: "Find out how good your tracking really is (confidence, gaps, and whether the body keeps its shape) before changing anything."
order: 3
---

```{r}
#| label: setup
#| include: false
library(animovement)
knitr::opts_chunk$set(
  comment = "#>",
  fig.width = 7, fig.height = 4.5, fig.align = "center",
  dpi = 150, out.width = "100%"
)
```

Tracking data is never perfect. Keypoints get occluded, the tracker guesses,
and once a guess is written to file it looks exactly like a measurement: a pair
of plausible coordinates.

[anicheck](https://animovement.dev/anicheck) helps you tell them apart. Its
functions only *report*. Each takes an anipoint and returns a summary, leaving
your data untouched. Cleaning comes in the next guide, once you know what
needs cleaning.

We start from where the [previous guide](read-data.qmd) left off:

```{r}
#| label: read
library(animovement)

path <- get_sample_data("sleap")
movement <- read_dataset(path) |>
  set_metadata(sampling_rate = 30) |>
  convert_unit_time("s") |>
  convert_unit_space("cm", calibration_factor = 9 / 140) |>
  set_structure(read_structure(path))
```

Most checks report on every keypoint at once. When we look closer, we'll follow
one: the **centre** of the mouse, which is the point most analyses of where an
animal goes and how fast are based on.

## How confident was the tracker? {#sec-confidence}

Pose-estimation tools score every point they place. `check_confidence()`
summarises those scores per keypoint:

```{r}
#| label: check-conf
check_confidence(movement)
```

The centre is tracked most consistently, while the snout and the tail end are
tracked far less confidently. That makes sense: they're the parts that
disappear when the mouse turns. To see the whole distribution rather than a
summary, plot the check:

```{r}
#| label: plot-conf
#| fig-cap: "Confidence per keypoint. The spike at zero is the tracker admitting it had no idea."
plot(check_confidence(movement))
```

This is the plot to choose a threshold from. Most of each keypoint's scores sit
near 1, with a separate clump at zero and a thin spread in between, so a cut
around 0.6 removes the failures without touching the good tracking. Where
exactly to cut is a judgement about your data, and this plot lets you make it
rather than guess.

::: {.callout-note}
## Confidence isn't a probability
Scores are whatever the tracker chose to emit. SLEAP's run past 1 in this
recording, and a 0.9 from one tool says nothing about a 0.9 from another. Read
the distribution, rather than borrowing a threshold from someone else's paper.
:::

## Where are the gaps? {#sec-gaps}

Missing values are the other half of the story. `check_na_gapsize()` reports how
many there are and, more usefully, whether they come as short dropouts or long
absences:

```{r}
#| label: check-gaps
check_na_gapsize(movement)
```

The overall percentage hides the shape of the problem, and it differs a lot
between keypoints. The snout and the tail end have hundreds of gaps each. The
centre has far fewer, though its longest still runs to over a hundred frames.
Plotting the centre on its own shows the shape:

```{r}
#| label: plot-gaps
#| fig-height: 3.5
#| fig-cap: "Gap sizes for the centre. Most gaps are a handful of frames long."
movement |>
  dplyr::filter(keypoint == "centre") |>
  check_na_gapsize() |>
  plot()
```

The difference matters for the next step. A gap of one frame is worth
interpolating across, because the mouse can't have gone far. A gap of a hundred
frames is over three seconds, and anything drawn across it is invention.

::: {.callout-tip}
## Plotting several groups at once
These check plots put each group in its own panel, so with six keypoints you'll
want a taller figure (`fig-height: 10` or so) before they become readable. One
keypoint at a time is usually easier to read anyway.
:::

## When do they happen? {#sec-timing}

Gaps clustered in one stretch mean something different from gaps spread evenly:
a lighting change, say, rather than a keypoint the model was never good at.
`check_na_timing()` splits the recording into intervals and shows how much of
each is missing:

```{r}
#| label: check-timing
#| fig-height: 3.5
#| fig-cap: "When the centre goes missing, interval by interval."
movement |>
  dplyr::filter(keypoint == "centre") |>
  check_na_timing() |>
  plot()
```

Very little of the centre is missing, with two worse patches: at the very
start, and around seven and a half minutes in. Neither is a broken stretch of
recording worth throwing away. They're more likely moments when the mouse is
hard to see, which is worth remembering when you interpret what it did then.

## Does the body keep its shape? {#sec-segments}

A mouse can't stretch to twice its length between two frames, so a body segment
whose length jumps around can point to a keypoint placed in the wrong spot.
With the skeleton attached, `check_segment_length()` compares each segment's
length in every frame with its usual length:

```{r}
#| label: check-segments
check_segment_length(movement)
```

```{r}
#| label: plot-segments
#| fig-height: 5
#| fig-cap: "Each segment's length relative to its usual length. The dashed lines mark 30% either way."
plot(check_segment_length(movement))
```

Read this with care in 2D. A camera looking down sees each segment's length
*projected* onto the floor, and that changes whenever the mouse rears, lowers
its head or turns it. So segments vary in length even when the tracking is
perfect. The short segments from the snout to the ears vary most, because a
small movement of the head changes them by a large fraction. In 3D, where the
true length is measured, the same check is much stricter. In 2D it is better at
catching gross errors, such as two keypoints swapped, than at judging precision.

## What we learned {#sec-summary}

What the checks mean for the [next guide](process-data.qmd):

- The maze occupies roughly x from 19 to 65 cm and y from 0 to 45 cm. Points
  outside it are impossible, so they can go.
- Confidence separates cleanly around **0.6**, so that's where we'll mask.
- Most gaps are a few frames long and safe to interpolate, but some run to
  several seconds and are not.
- No stretch of the recording is broken badly enough to discard.

Next: [process the trajectories](process-data.qmd).
