---
title: "Install animovement"
subtitle: "Install the animovement ecosystem in R or with Pixi, load it, and add the optional packages."
order: 1
code-annotations: hover
---


```{r}
#| label: setup
#| include: false
# animovement is deliberately NOT attached here; the visible chunk below does
# it, so the page shows the real attach banner.
knitr::opts_chunk$set(comment = "#>")
```

## Install the ecosystem {#sec-install}

Installing the `animovement` metapackage pulls in the whole ecosystem. There are
two ways to get it: from R itself, or as conda packages with
[Pixi](https://pixi.sh).

::: {.panel-tabset}
## R

*animovement* is distributed through
[our r-universe](https://animovement.r-universe.dev):

```{r}
#| eval: false
install.packages(
  "animovement",
  repos = c(
    "https://animovement.r-universe.dev",
    "https://cloud.r-project.org" # <1>
  )
)
```

1. The second repository (CRAN) is there so that dependencies which don't live on r-universe are found too. Once the metapackage is installed you can get that repository pair back at any time with `animovement_repos()`, which is what the other install helpers use by default.

## Pixi

Every package is also published as a conda package, on the
[animovement channel](https://prefix.dev/channels/animovement) on prefix.dev.
With [Pixi](https://pixi.sh), create a project that uses that channel alongside
conda-forge, add the metapackage, and start R inside it:

```bash
pixi init my-analysis \
  --channel https://prefix.dev/animovement \
  --channel conda-forge \
  --channel bioconda
cd my-analysis
pixi add r-animovement bioconductor-rhdf5
pixi run R
```

The project pins R and every package version in its lockfile, so the same
analysis installs identically on another machine. The conda packages are built
from the same commits as r-universe; `pixi update` brings them up to date.

`bioconductor-rhdf5`, from the bioconda channel, is the optional package the
SLEAP and DeepLabCut HDF5 readers need ([see below](#sec-suggested)). In a Pixi
project, add optional packages like this, with `pixi add`, rather than from
inside R.

::: {.callout-warning}
## Already have R installed?
R also looks in your personal package library, and it looks there first. If
that library holds packages built for the same R version, the project's R can
load those instead of its own and crash. To keep the project to its own
packages, add this to `pixi.toml`:

```toml
[activation.env]
R_LIBS_USER = "$CONDA_PREFIX/lib/R/library"
```
:::
:::

::: {.callout-tip}
## Only need one piece?
Every package installs the same way. Swap `"animovement"` for the package you
want, such as `"aniread"`, or with Pixi add `r-aniread`. See
[Packages](../packages/index.qmd) for what each one does.
:::

## Load it {#sec-load}

Like the tidyverse, `library(animovement)` attaches the core packages in one
call and tells you what you got:

```{r}
#| label: attach
library(animovement)
```

That's the metapackage plus the `r length(animovement_packages()) - 1` working
packages, one per stage of the workflow:

```{r}
#| label: show-packages
animovement_packages()
```

From here on, the guides assume you've run `library(animovement)`.

## Optional packages {#sec-suggested}

Some functions need a package that isn't installed by default, such as `arrow`
for Parquet files or `rhdf5` for HDF5 exports. To see which:

```{r}
#| label: suggested
animovement_show_suggested()
```

You don't have to install these up front. When a function needs one, it says
so and offers to install it there and then:

```r
read_dataset(get_sample_data("sleap"))
#> Downloading sleap data...
#> ℹ The package "rhdf5" is required for reading HDF5 files,
#> ✖ Would you like to install it?
#>
#> 1: Yes
#> 2: No
```

Answer `1` and it installs and carries on with what you asked for, without
restarting or re-running anything. To get them all in one go instead:

```{r}
#| eval: false
animovement_install_suggested()
```

::: {.callout-note}
## HDF5 exports need `rhdf5`
The sample used in these guides is a SLEAP HDF5 file, so `rhdf5` is the
optional package you're most likely to meet first. It comes from Bioconductor
rather than CRAN, so `install.packages("rhdf5")` won't find it. Use the prompt
above or `animovement_install_suggested()`, or with Pixi,
`pixi add bioconductor-rhdf5`.
:::

## Update packages {#sec-update}

The ecosystem is under active development. To update every *animovement*
package at once:

```{r}
#| eval: false
animovement_update()
```

In a Pixi project, run `pixi update` instead.
