---
title: "Install animovement"
subtitle: "Install the animovement ecosystem from r-universe, load it, and add the optional packages."
order: 1
code-annotations: hover
---


```{r}
#| label: setup
#| include: false
# animovement is deliberately NOT attached here — the visible chunk below does
# it, so the page shows the real attach banner.
knitr::opts_chunk$set(comment = "#>")
```

## Install the ecosystem {#sec-install}

*animovement* is distributed through
[our r-universe](https://animovement.r-universe.dev). Installing the `animovement`
metapackage pulls in the whole ecosystem:

```{r}
#| eval: false
install.packages(
  "animovement",
  repos = c(
    "https://animovement.r-universe.dev",
    "https://cloud.r-project.org" # <1>
  )
)
```

1. The second repository (CRAN) is there so that dependencies which don't live on r-universe are found too. Once the metapackage is installed you can get that repository pair back at any time with `animovement_repos()`, which is what the other install helpers use by default.

::: {.callout-tip}
## Only need one piece?
Every package installs the same way — just swap `"animovement"` for the
individual package name, for example `"aniread"`. See
[Packages](../packages/index.qmd) for what each one does.
:::

## Load it {#sec-load}

Like the tidyverse, `library(animovement)` attaches the core packages in one
call and tells you what you got:

```{r}
#| label: attach
library(animovement)
```

That's the metapackage plus the `r length(animovement_packages()) - 1` working
packages, one per stage of the workflow:

```{r}
#| label: show-packages
animovement_packages()
```

From here on, the guides assume you've run `library(animovement)`.

## Optional packages {#sec-suggested}

Some functions need a package that isn't installed by default. See which packages are suggested with:

```{r}
#| label: suggested
animovement_show_suggested()
```

You don't have to install these up front. When a function needs one it will
say so and offer to install it there and then:

```r
read_sleap(get_sample_data("sleap"))
#> Downloading sleap data...
#> ℹ The package "rhdf5" is required for reading HDF5 files,
#> ✖ Would you like to install it?
#>
#> 1: Yes
#> 2: No
```

Answer `1` and it installs and carries on with what you asked for without needing to
restart or re-run. To get them all in one go instead:

```{r}
#| eval: false
animovement_install_suggested()
```

::: {.callout-note}
## HDF5 exports need `rhdf5`
The samples used in these guides are SLEAP and DeepLabCut HDF5 files, so
`rhdf5` is the one you're most likely to meet first. It comes from
Bioconductor rather than CRAN, so `install.packages("rhdf5")` won't find it —
use the prompt above when you encounter it, or `animovement_install_suggested()`.
:::

## Update packages {#sec-update}

The ecosystem is under active development. To update every *animovement*
package at once:

```{r}
#| eval: false
animovement_update()
```
