---
title: "Process trajectories"
description: "Mask the points you don't trust, fill the gaps worth filling, and smooth out the jitter."
order: 4
---

```{r}
#| label: setup
#| include: false
library(animovement)
knitr::opts_chunk$set(
  comment = "#>",
  fig.width = 7, fig.height = 4.5, fig.align = "center",
  dpi = 150, out.width = "100%"
)
```

You rarely want to analyse completely raw tracking. There are points you don't
trust, gaps worth filling, and jitter to smooth out.
[aniprocess](https://animovement.dev/aniprocess) handles all three, and each
has its own family of functions:

1. **Mask** the points you don't trust, turning them into honest `NA`s, with
   `mask_na_*()`.
2. **Fill** the gaps short enough to interpolate across, with `replace_na_*()`.
3. **Smooth** what's left to take out frame-to-frame jitter, with `filter_*()`.

Do them in that order. Masking first turns bad points into gaps, so that the
filling step bridges them instead of building on them, and smoothing last works
on a complete, trustworthy track.

```{r}
#| label: read
library(animovement)

movement <- read_dataset(get_sample_data("sleap")) |>
  set_metadata(sampling_rate = 30) |>
  convert_unit_time("s")
```

Each family has a function ending in `_across()` that works on a whole anipoint:
it applies a method to every position column, within each track, using the
frame's metadata to find them. That's what we use here.

## 1. Mask what you don't trust {#sec-mask}

Two criteria earn their place for this recording, and they catch different
kinds of error.

**Where the animal cannot have been.** The maze runs from roughly x = 300 to
1010 and y = 0 to 690, as you can read off the
[first plot](read-data.qmd#sec-look). Anything outside that box is not the
mouse:

```{r}
#| label: mask-roi
in_maze <- movement |>
  mask_na_across("roi", x_min = 300, x_max = 1010, y_min = 0, y_max = 690)
```

A rectangle around a plus-shaped maze only rules out the impossible, not
everything implausible, since the corners of the box are floor the mouse can't
reach either. But it costs nothing, and it removes every stray point.

**Where the tracker wasn't sure.** The [previous guide](check-data.qmd) put the
confidence threshold at 0.6:

```{r}
#| label: mask-confidence
masked <- in_maze |>
  mask_na_across("confidence", threshold = 0.6)
```

The warning is worth reading: some rows have no confidence value at all, so
there's nothing to compare against the threshold, and they're left alone.

Between them, the two masks bring the missing data to
`r round(100 * mean(is.na(masked$x)), 1)`%. What was guesswork is now visible
as absence.

Using two criteria is normal, not overcautious. A confidence threshold alone
would have kept the stray points, because the tracker was confident about them.
A region alone would keep every low-confidence guess that happens to land on
the maze. Being wrong *in space* and being wrong *by the tracker's own
estimate* are different failures, and each needs its own test.

`mask_na_across()` has more criteria to choose from:

| Method | Masks |
|--------|-------|
| `"range"` | values outside plausible bounds |
| `"roi"` | points outside a region of interest |
| `"confidence"` | points scored below a threshold |
| `"speed"` | jumps too fast to be real |
| `"excursion"` | a point that leaps away and later comes back |
| `"hampel"` | single-frame spikes, judged against the points around them |

With a skeleton attached, `mask_na_segment_length()` masks points that stretch
a body segment far beyond its usual length.

## 2. Fill the gaps worth filling {#sec-fill}

Now interpolate, but only across gaps short enough to be honest about. `max_gap`
is the safety valve: longer gaps are left missing rather than filled with an
invented trajectory.

```{r}
#| label: fill
filled <- masked |>
  replace_na_across("linear", max_gap = 5)
```

Five frames is a sixth of a second here. Gaps that short get a straight line
across them, and the long absences found earlier stay `NA`. That's why
`r round(100 * mean(is.na(filled$x)), 1)`% is still missing, as it should be.

Other methods swap in by name: `"spline"` for a smoother curve through the gap,
`"stine"` for Stineman interpolation, `"locf"` to carry the last value forward,
and `"value"` for a constant.

## 3. Smooth the jitter {#sec-smooth}

Even well-tracked keypoints wobble a pixel or two from frame to frame, and that
wobble becomes noise as soon as you differentiate position into speed. The
Savitzky–Golay filter fits a local polynomial, so it keeps the shape of real
movement better than a plain moving average:

```{r}
#| label: smooth
clean <- filled |>
  filter_across("sgolay", window_width = 11, order = 3)
```

A window of 11 frames at 30 fps is about a third of a second. Widen it and the
track gets smoother, but real detail starts to disappear. For a mouse's snout,
this is roughly the limit. Other methods include `"rollmedian"`, which resists
spikes, `"gaussian"` and `"triangular"` for gentle weighted smoothing, and
`"lowpass"` and `"kalman"` for more control.

## Did it work? {#sec-verify}

Run the checks again. That's what they're for:

```{r}
#| label: recheck
check_na_gapsize(clean)
```

And look at it. Here are fifteen seconds of the snout, before and after:

```{r}
#| label: before-after
#| fig-width: 10
#| fig-height: 4
#| fig-cap: "The snout over fifteen seconds: raw on the left, processed on the right."
snout_window <- function(data) {
  dplyr::filter(data, keypoint == "snout", time > 100, time < 115)
}

plots(
  plot(snout_window(movement)),
  plot(snout_window(clean))
)
```

The processed track follows the same path, with the jitter gone and the short
dropouts bridged. The long gaps are left as holes on purpose.

## The whole thing {#sec-together}

Three steps in one pipe, and the result is still an anipoint:

```{r}
#| label: pipeline
#| eval: false
clean <- read_dataset(get_sample_data("sleap")) |>
  set_metadata(sampling_rate = 30) |>
  convert_unit_time("s") |>
  mask_na_across("roi", x_min = 300, x_max = 1010, y_min = 0, y_max = 690) |>
  mask_na_across("confidence", threshold = 0.6) |>
  replace_na_across("linear", max_gap = 5) |>
  filter_across("sgolay", window_width = 11, order = 3)
```

::: {.callout-note}
## `filter_*()` doesn't drop rows
In *animovement*, `filter_` means **signal filtering**: `filter_across()`
smooths, and nothing in aniprocess drops a row. `dplyr::filter()` still works on
an anipoint and still subsets rows, as it does on any data frame.
:::

Next: [calculate metrics](calculate-metrics.qmd).
