---
title: "Process trajectories"
subtitle: "You rarely want completely raw movement data: there may be observations you don't trust, gaps that are worth filling, and jitter to smooth out. Here we give you a brief overview of how to do this in `animovement`."
order: 4
---

```{r}
#| label: setup
#| include: false
library(animovement)
knitr::opts_chunk$set(
  comment = "#>",
  fig.width = 7, fig.height = 4.5, fig.align = "center",
  dpi = 150, out.width = "100%"
)
```

In *animovement*, [aniprocess](https://animovement.dev/aniprocess) is the package responsible for the cleaning. Here you have three possible actions

1. **Mask** the points you don't trust, turning them into honest `NA`s (`filter_na_*`).
2. **Fill** the gaps short enough to interpolate across (`replace_na_*`).
3. **Smooth** what's left, to take out frame-to-frame jitter (`filter_*`).

It is generally good practice to approach the steps in that order, masking first (turning unreliable observations into gaps), before filling gaps and finally smoothing the trajectories which can produce more realistic tracks.

```{r}
#| label: read
library(animovement)

movement <- read_sleap(get_sample_data("sleap")) |>
  set_sampling_rate(30)
```

::: {.callout-warning}
## `filter_*()` doesn't filter rows
In *animovement*, `filter_` means **signal filtering**. `filter_sgolay()`
smooths, `filter_na_confidence()` masks bad points to `NA` — neither one drops
a row. `dplyr::filter()` still works on an aniframe and still subsets rows, as
it always did. Two different jobs, unfortunately similar names.
:::

## 1. Mask what you don't trust {#sec-mask}

`filter_na_across()` applies a criterion across every spatial column at once,
using the aniframe's own metadata to know which those are. Two criteria earn
their place here, and they catch different kinds of wrong.

**Where the animal cannot have been.** The maze runs from roughly x = 300 to
1010 and y = 0 to 690 — read straight off the [first plot](read-data.qmd#sec-look).
Anything outside that box is not a mouse:

```{r}
#| label: mask-roi
in_maze <- movement |>
  filter_na_across("roi", x_min = 300, x_max = 1010, y_min = 0, y_max = 690)
```

That's a rectangle around a plus-shaped maze, so it only rules out the
impossible, not everything implausible — the corners of the box are floor the
mouse can't reach either. It costs almost nothing and removes every stray point:
missing data goes from 8.3% to just
`r round(100 * mean(is.na(in_maze$x)), 2)`%, and the excursions are gone.

**Where the tracker wasn't sure.** The [previous guide](check-data.qmd) put the
confidence threshold at 0.6:

```{r}
#| label: mask
masked <- in_maze |>
  filter_na_across("confidence", threshold = 0.6)
```

The warning is worth reading rather than dismissing: some rows have no
confidence value at all, so there's nothing to compare against a threshold and
they're left alone. Between them the two masks take the missing data to
`r round(100 * mean(is.na(masked$x)), 1)`% — the guesswork, now visible as
absence:

```{r}
#| label: mask-pct
round(100 * mean(is.na(masked$x)), 1)
```

Confidence is one criterion among several. `filter_na_across()` also takes
`"range"` (values outside plausible bounds), `"roi"` (outside a region of
interest), `"speed"` (jumps too fast to be real) and `"excursion"` (a point that
leaps away and comes straight back). Each has a standalone form too —
`filter_na_speed()` and friends — for when you want to treat one column
differently from the rest.

Using two criteria together is the norm, not belt-and-braces. A confidence
threshold alone would have left the stray points in place — the tracker was
perfectly confident about them — and an ROI alone keeps every low-confidence
guess that happens to land on the maze. Wrong *in space* and wrong *in the
tracker's own estimation* are different failures, and each needs its own test.

## 2. Fill the gaps worth filling {#sec-fill}

Now interpolate, but only across gaps short enough to be honest about. `max_gap`
is the safety valve — leave the long absences missing rather than inventing a
trajectory through them:

```{r}
#| label: fill
filled <- masked |>
  replace_na_across("linear", max_gap = 5)

round(100 * mean(is.na(filled$x)), 1)
```

Five frames is a sixth of a second here. Gaps that short get a straight line
drawn across them; the 197-frame absences we found earlier stay `NA`, which is
why `r round(100 * mean(is.na(filled$x)), 1)`% is still missing — and should be.

Other methods swap in by name: `"spline"` for a smoother curve through the gap,
`"stine"` for Stineman interpolation, `"locf"` to carry the last value forward,
`"value"` for a constant.

## 3. Smooth the jitter {#sec-smooth}

Even well-tracked keypoints wobble a pixel or two per frame, and that wobble
becomes noise the moment you differentiate it into speed.
`filter_across("sgolay")` applies a Savitzky–Golay filter, which fits a local
polynomial and so preserves the shape of real movement better than a plain
moving average:

```{r}
#| label: smooth
clean <- filled |>
  filter_across("sgolay", window_width = 11, order = 3)
```

A window of 11 frames at 30 fps is about a third of a second. Widen it and the
track gets smoother but real detail starts to disappear — for a mouse's snout,
this is roughly the limit. Alternatives: `"rollmedian"` (spike-resistant, keeps
edges crisp), `"gaussian"` or `"triangular"` (gentle weighted smooths),
`"lowpass"` and `"kalman"` for more control.

## Did it work? {#sec-verify}

Run the same checks as before — that's the point of them being cheap:

```{r}
#| label: recheck
check_na_gapsize(clean)
```

And look at it. Fifteen seconds of the snout, before and after:

```{r}
#| label: before-after
#| fig-width: 10
#| fig-height: 4
#| fig-cap: "The snout over fifteen seconds: raw on the left, processed on the right."
window <- function(data) {
  dplyr::filter(data, keypoint == "snout", time > 100, time < 115)
}

plots(
  plot(window(movement)),
  plot(window(clean))
)
```

The processed track keeps the same path, with the jitter gone and the short
dropouts bridged. The dashed segments are the gaps that were too long to fill —
left as holes on purpose.

## The whole thing {#sec-together}

Three steps, one pipe, still an aniframe at the end:

```{r}
#| label: pipeline
#| eval: false
clean <- read_sleap(get_sample_data("sleap")) |>
  set_sampling_rate(30) |>
  filter_na_across("roi", x_min = 300, x_max = 1010, y_min = 0, y_max = 690) |>
  filter_na_across("confidence", threshold = 0.6) |>
  replace_na_across("linear", max_gap = 5) |>
  filter_across("sgolay", window_width = 11, order = 3)
```

Next: [calculate metrics](calculate-metrics.qmd).
