---
title: "Process trajectories"
description: "Mask the points you don't trust, fill the gaps worth filling, and smooth out the jitter, checking each step as you go."
order: 4
---

```{r}
#| label: setup
#| include: false
library(animovement)
knitr::opts_chunk$set(
  comment = "#>",
  fig.width = 7, fig.height = 4.5, fig.align = "center",
  dpi = 150, out.width = "100%"
)
```

You rarely want to analyse completely raw tracking. There are points you don't
trust, gaps worth filling, and jitter to smooth out.
[aniprocess](https://animovement.dev/aniprocess) handles all three, and each
has its own family of functions:

- **Mask** the points you don't trust, turning them into honest `NA`s, with
  `mask_na_*()`.
- **Fill** the gaps short enough to interpolate across, with `replace_na_*()`.
- **Smooth** what's left to take out frame-to-frame jitter, with `filter_*()`.

Do them in that order. Masking first turns bad points into gaps, so that the
filling step bridges them instead of building on them, and smoothing last works
on a complete, trustworthy track. After each step, run a check: the
[checks](check-data.qmd) are cheap, and they show you what the step actually
did to your data.

```{r}
#| label: read
library(animovement)

movement <- read_dataset(get_sample_data("sleap")) |>
  set_metadata(sampling_rate = 30) |>
  convert_unit_time("s") |>
  convert_unit_space("cm", calibration_factor = 9 / 140)
```

Each family has a function ending in `_across()` that works on a whole anipoint:
it applies a method to every position column, within each track, using the
frame's metadata to find them. That's what we use here.

## Mask what you don't trust {#sec-mask}

Two criteria earn their place for this recording, and they catch different
kinds of error.

**Where the animal cannot have been.** The maze runs from roughly x = 19 to
65 cm and y = 0 to 45 cm, as you can read off the
[first plot](read-data.qmd#sec-look). Anything outside that box is not the
mouse:

```{r}
#| label: mask-roi
in_maze <- movement |>
  mask_na_across("roi", x_min = 19, x_max = 65, y_min = 0, y_max = 45)
```

A rectangle around a plus-shaped maze only rules out the impossible, not
everything implausible, since the corners of the box are floor the mouse can't
reach either. But it costs nothing, and it removes every stray point.

**Where the tracker wasn't sure.** The [previous guide](check-data.qmd) put the
confidence threshold at 0.6:

```{r}
#| label: mask-confidence
masked <- in_maze |>
  mask_na_across("confidence", threshold = 0.6)
```

Rows with no confidence score at all are left alone, since a missing score
means *not assessed* rather than *poor*. Set `missing = "mask"` to mask them
too. In this recording those rows have no position anyway.

Check what the masks did. The confidence distribution now stops at 0.6, and
the clump of guesses at zero is gone:

```{r}
#| label: check-masked
#| fig-cap: "Confidence per keypoint after masking."
plot(check_confidence(masked))
```

They also turned those guesses into gaps:

```{r}
#| label: gaps-masked
check_na_gapsize(masked)
```

Missing data has gone from `r round(100 * mean(is.na(movement$x)), 1)`% to
`r round(100 * mean(is.na(masked$x)), 1)`%. Every keypoint has more gaps than
before, and the snout, the tail end and the right ear, whose confidence was
lowest, gained the most. What was guesswork is now visible as absence.

Using two criteria is normal, not overcautious. A confidence threshold alone
would have kept the stray points, because the tracker was confident about them.
A region alone would keep every low-confidence guess that happens to land on
the maze. Being wrong *in space* and being wrong *by the tracker's own
estimate* are different failures, and each needs its own test.

`mask_na_across()` has more criteria to choose from:

| Method | Masks |
|--------|-------|
| `"range"` | values outside plausible bounds |
| `"roi"` | points outside a region of interest |
| `"confidence"` | points scored below a threshold |
| `"speed"` | jumps too fast to be real |
| `"excursion"` | a point that leaps away and later comes back |
| `"hampel"` | single-frame spikes, judged against the points around them |

## Fill the gaps worth filling {#sec-fill}

Now interpolate, but only across gaps short enough to be honest about. `max_gap`
is the safety valve: longer gaps are left missing rather than filled with an
invented trajectory.

```{r}
#| label: fill
filled <- masked |>
  replace_na_across("linear", max_gap = 5)
```

Five frames is a sixth of a second here. Check the centre's gaps again, before
and after filling:

```{r}
#| label: check-before-fill
#| fig-height: 3.5
#| fig-cap: "Gap sizes for the centre before filling."
centre <- function(data) dplyr::filter(data, keypoint == "centre")

plot(check_na_gapsize(centre(masked)))
```

```{r}
#| label: check-after-fill
#| fig-height: 3.5
#| fig-cap: "Gap sizes for the centre after filling. Only the gaps longer than five frames remain."
plot(check_na_gapsize(centre(filled)))
```

The short gaps are gone and the long ones are untouched. That's why
`r round(100 * mean(is.na(filled$x)), 1)`% of the data is still missing, as it
should be.

Other methods swap in by name: `"spline"` for a smoother curve through the gap,
`"stine"` for Stineman interpolation, `"locf"` to carry the last value forward,
and `"value"` for a constant.

## Smooth the jitter {#sec-smooth}

Even well-tracked keypoints wobble a little from frame to frame, and that
wobble becomes noise as soon as you differentiate position into speed. The
Savitzky–Golay filter fits a local polynomial, so it keeps the shape of real
movement better than a plain moving average:

```{r}
#| label: smooth
clean <- filled |>
  filter_across("sgolay", window_width = 11, order = 3)
```

A window of 11 frames at 30 fps is about a third of a second. Widen it and the
track gets smoother, but real detail starts to disappear. Other methods include
`"rollmedian"`, which resists spikes, `"gaussian"` and `"triangular"` for gentle
weighted smoothing, and `"lowpass"` and `"kalman"` for more control.

Smoothing doesn't change what's missing, so here the check to run is your eyes.
Here are fifteen seconds of the centre, as the mouse runs along an arm, through
the middle of the maze, up the opposite arm and back:

```{r}
#| label: before-after
#| fig-width: 10
#| fig-height: 4
#| fig-cap: "Fifteen seconds of the mouse's centre: raw on the left, processed on the right."
run <- function(data) dplyr::filter(centre(data), time > 335, time < 350)

plots(
  plot(run(movement)),
  plot(run(clean))
)
```

The processed track follows the same path, with the jitter gone and the short
dropouts bridged.

## The whole thing {#sec-together}

Three steps in one pipe, and the result is still an anipoint:

```{r}
#| label: pipeline
#| eval: false
clean <- read_dataset(get_sample_data("sleap")) |>
  set_metadata(sampling_rate = 30) |>
  convert_unit_time("s") |>
  convert_unit_space("cm", calibration_factor = 9 / 140) |>
  mask_na_across("roi", x_min = 19, x_max = 65, y_min = 0, y_max = 45) |>
  mask_na_across("confidence", threshold = 0.6) |>
  replace_na_across("linear", max_gap = 5) |>
  filter_across("sgolay", window_width = 11, order = 3)
```

::: {.callout-note}
## `filter_*()` doesn't drop rows
In *animovement*, `filter_` means **signal filtering**: `filter_across()`
smooths, and nothing in aniprocess drops a row. `dplyr::filter()` still works on
an anipoint and still subsets rows, as it does on any data frame.
:::

Next: [calculate metrics](calculate-metrics.qmd).
