---
title: "Read your data"
description: "Turn a tracking export into an anipoint with aniread, and tell it about your recording."
order: 2
---

```{r}
#| label: setup
#| include: false
library(animovement)
knitr::opts_chunk$set(
  comment = "#>",
  fig.width = 7, fig.height = 4.5, fig.align = "center",
  dpi = 150, out.width = "100%"
)
```

[aniread](https://animovement.dev/aniread) reads movement data from a wide range
of sources: pose estimation from video, centroid trackers, motion capture, even
trackball hardware. Whichever tool wrote the file, you get back the same
structure, so after this first step it stops mattering.

## Get some data to follow along {#sec-sample}

You don't need your own files. `get_sample_data()` downloads a real export for
any supported source and returns its path:

```{r}
#| label: sample
path <- get_sample_data("sleap")
basename(path)
```

We'll use this recording in all of the guides: ten minutes of a single mouse on
an elevated plus maze, tracked with [SLEAP](https://sleap.ai). To see what else
is available, call `get_sample_data(list_datasets = TRUE)`.

## Read it {#sec-read}

`read_dataset()` reads any supported file. It works out which tool wrote it, so
you don't need to know which reader it needs:

```{r}
#| label: read
movement <- read_dataset(path)
movement
```

## What you got back {#sec-anipoint}

That's an **anipoint**: one row per point (here, each keypoint of each
individual) at each moment. It's one of the *aniframe* classes defined by
[anicore](https://animovement.dev/anicore), which every other package in the
ecosystem reads and returns. Underneath it's a
[tibble](https://tibble.tidyverse.org), so `dplyr` and `ggplot2` work on it as
usual. On top, metadata records what each column is for:

| Question | Role | Read it with | Here |
|----------|------|--------------|------|
| **What** is moving? | identity | `get_variables(movement, "what")` | `individual`, `keypoint` |
| **When**? | index | `get_index(movement)` | `time` |
| **Where**? | position | `get_axes(movement)` | `x`, `y` |

```{r}
#| label: roles
get_variables(movement, "what")
get_index(movement)
get_axes(movement)
```

Two consequences worth knowing from the start:

- **Column names are free.** The roles are what matter, so a file whose identity
  column is called `track` works just as well.
- **The frame arrives grouped by identity**, so everything downstream runs
  within each track rather than across them:

```{r}
#| label: groups
dplyr::n_groups(movement)
```

Six groups: one mouse × six keypoints. That grouping is why interpolation
never bleeds from the snout into the tail, and why speed is never computed
across the join between two tracks.

::: {.callout-tip}
## Going deeper
This is the short version. anicore's articles cover the data structure in full:
[The aniframe data structure](https://animovement.dev/anicore/articles/aniframe-structure.html)
for the roles and how to set them, and
[Metadata on an aniframe](https://animovement.dev/anicore/articles/aniframe-metadata.html)
for everything below.
:::

## Take a first look {#sec-look}

Calling `plot()` on an anipoint draws the trajectories, with axis labels taken
from the metadata:

```{r}
#| label: first-plot
#| fig-cap: "Raw SLEAP tracks. The plus maze is clearly visible, and so are a handful of points that landed nowhere near it."
plot(movement)
```

The maze shape comes straight out of the data, with no annotation needed. So do
the long lines reaching off to the left: a few dozen points, out of
`r format(nrow(movement), big.mark = ",")`, that landed well outside the maze.
They're drawn as long straight lines because the plot joins consecutive
positions. Whether that matters, and what else is hiding in there, is what the
[next guide](check-data.qmd) is for.

## Tell it about your recording {#sec-metadata}

An anipoint keeps context that a bare data frame would lose. The reader fills in
what the file records:

```{r}
#| label: metadata
get_metadata(movement)
```

Note `sampling_rate`. SLEAP's analysis files don't record a frame rate, so `time`
counts frames, and nothing downstream can turn frames into seconds until you
say what the rate was. Declare it, then convert the time column to seconds:

```{r}
#| label: set-rate
movement <- movement |>
  set_metadata(sampling_rate = 30) |>
  convert_unit_time("s")
```

The printed header now shows the rate and the duration, and from here on speeds
come out per second rather than per frame:

```{r}
#| label: after-rate
movement
```

::: {.callout-warning}
## Use your own frame rate
We're assuming 30 fps for this sample. Set whatever your recording actually
was: everything downstream that involves time scales with this number.
:::

Coordinates are in pixels (`unit_space: px`), which is the honest answer for a
video export. If you've measured how many pixels make up a known distance,
`convert_unit_space()` converts the coordinates, and distances and speeds come
out in real-world units. Sampling rate, axis directions and handedness are set
the same way, with `set_metadata()`, and read back with `get_metadata()`.

## Attach the skeleton {#sec-structure}

Pose-estimation tools also record which keypoints are joined to which: the
skeleton. `read_structure()` reads it from the same file, and `set_structure()`
attaches it to the frame:

```{r}
#| label: structure
movement <- movement |>
  set_structure(read_structure(path))

get_structure(movement)
```

With the skeleton attached, functions that work on body segments know what the
segments are. The [next guide](check-data.qmd) uses it to check whether the
mouse keeps its shape. The anicore article on
[structures](https://animovement.dev/anicore/articles/structures.html) covers
them in full.

## Reading your own files {#sec-own}

`read_dataset()` is the one function to remember. Point it at a file, and it
detects the source and calls the matching reader. To see what's supported:

```{r}
#| label: sources
get_supported_sources()
```

If you're curious what it decides a file is, `detect_source()` tells you
without reading the file:

```{r}
#| label: detect
detect_source(path)
```

When detection can't tell two sources apart, or you want to be explicit, name
the source: `read_dataset(path, source = "sleap")`. Each source also has its own
reader, such as `read_sleap()` or `read_deeplabcut()`, whose help pages document
the options specific to that format. Pass those options to `read_dataset()` and
it hands them on.

For a format aniread doesn't cover, `read_custom()` maps arbitrary columns onto
the roles above. And once you've read something, `write_aniframe()` saves it
to a Parquet file, metadata and all, which `read_aniframe()` opens again far
faster than re-parsing the original export.

Next: [check what you actually have](check-data.qmd).
