---
title: "Read your data"
description: "Turn a tracking export into an aniframe with aniread, and tell it about your recording."
order: 2
---

```{r}
#| label: setup
#| include: false
library(animovement)
knitr::opts_chunk$set(
  comment = "#>",
  fig.width = 7, fig.height = 4.5, fig.align = "center",
  dpi = 150, out.width = "100%"
)
```

[aniread](https://animovement.dev/aniread) reads movement data from a wide range
of sources — video-based pose estimation, centroid trackers, motion capture,
even hardware trackballs — and hands back the same structure every time. Which
tool wrote your file is a detail that stops mattering after this first step.

## Get some data to follow along {#sec-sample}

You don't need your own files. `get_sample_data()` downloads a real export for
any supported source and returns the path to it:

```{r}
#| label: sample
path <- get_sample_data("sleap")
basename(path)
```

We'll use this recording throughout the guides: ten minutes of a single mouse on
an elevated plus maze, tracked with [SLEAP](https://sleap.ai). To see what else
is on offer, call `get_sample_data(list_datasets = TRUE)`; sources with several
datasets take a `dataset` argument, and passing a URL downloads any file you
point it at.

## Read it {#sec-read}

Every source has a dedicated `read_*()` function. Give it the path:

```{r}
#| label: read
movement <- read_sleap(path)
movement
```

## What you got back {#sec-aniframe}

That's an **aniframe** — the data structure defined by
[anicore](https://animovement.dev/anicore) that every other package in the
ecosystem reads and returns. It's a [tibble](https://tibble.tidyverse.org)
underneath, so `dplyr` and `ggplot2` work on it as usual, with metadata on top
recording what each column means:

| Question | Slot | Here |
|----------|------|------|
| **What** is moving? | `variables_what` | `individual`, `keypoint` |
| **When**? | `variables_index` | `time` |
| **Where**? | `variables_where` | `x`, `y` |

```{r}
#| label: slots
get_variables_what(movement)
get_index(movement)
get_axes(movement)
```

Two consequences worth knowing from the start. Column *names* are free — the
roles are what matter, so a file whose identity column is called `track` works
just as well. And an aniframe arrives **grouped by identity**, so everything
downstream runs within each track rather than across them:

```{r}
#| label: groups
dplyr::n_groups(movement)
```

Six groups — one mouse × six keypoints. That grouping is why interpolation never
bleeds from the snout into the tail, and why speed is never computed across the
join between two individuals.

::: {.callout-tip}
## Going deeper
This is the short version. anicore's own articles cover the data structure in
full — [The aniframe data structure](https://animovement.dev/anicore/articles/aniframe-structure.html)
for the roles and how to set them, and
[Metadata on an aniframe](https://animovement.dev/anicore/articles/aniframe-metadata.html)
for everything below.
:::

## Take a first look {#sec-look}

Calling `plot()` on an aniframe draws the trajectories, and reads the axis
labels off the metadata:

```{r}
#| label: first-plot
#| fig-cap: "Raw SLEAP tracks. The plus maze is clearly visible — and so is a handful of points that landed nowhere near it."
plot(movement)
```

The maze shape comes straight out of the data — no annotation needed. So do the
two long lines reaching off to the left: a few dozen points, out of 110,910,
that landed well outside the maze, drawn as long straight lines because the plot
joins consecutive positions. Whether that matters, and what else is hiding in
there, is what the [next guide](check-data.qmd) is for.

## Tell it about your recording {#sec-metadata}

An aniframe remembers context a bare data frame would lose. Some of it the
reader fills in for you:

```{r}
#| label: metadata
get_metadata(movement)
```

Note `sampling_rate` — SLEAP exports don't record a frame rate, so aniread can't
either, and `time` is counted in frames. Nothing downstream can turn frames into
seconds until you say what the rate was:

```{r}
#| label: set-rate
movement <- movement |>
  set_sampling_rate(30)
```

That one call changes what the rest of the workflow means. The time column is now
in seconds, the printed header gains a duration, and speeds will come out per
second instead of per frame:

```{r}
#| label: after-rate
movement
```

::: {.callout-warning}
## Use your own frame rate
We're assuming 30 fps for this sample. Set whatever your recording actually was
— everything downstream that involves time scales with this number.
:::

Coordinates are in pixels (`unit_space: px`), which is the honest answer for a
video export. If you've calibrated against something of known size, tell the
frame with `set_unit_space()` and distances and speeds become real-world units.
Sampling rate, units, axis directions and handedness are all set the same way,
with `set_*()` and read back with `get_*()`.

## Reading your own files {#sec-own}

The reader name matches the source:

| Source | Function |
|--------|----------|
| SLEAP | `read_sleap()` |
| DeepLabCut | `read_deeplabcut()` |
| LightningPose | `read_lightningpose()` |
| idtracker.ai | `read_idtracker()` |
| TRex | `read_trex()` |
| Anipose (3-D) | `read_anipose()` |
| Trackball hardware | `read_trackball()` |
| BORIS annotations | `read_boris()` |

Swap the reader, keep the rest of your workflow identical — that's the whole
point of a shared data structure. For the full list of what's supported:

```{r}
#| label: sources
get_supported_sources()
```

Not sure what wrote a file? `detect_source()` works it out, and `read_dataset()`
goes one further and reads it with whatever reader that turns out to be:

```{r}
#| label: detect
detect_source(path)
```

Have a format aniread doesn't cover? `read_custom()` maps arbitrary columns onto
the aniframe roles. And once you've read something, `write_aniframe()` saves it
— metadata and all — to a Parquet file that `read_aniframe()` opens again far
faster than re-parsing the original export.

Next: [check what you actually have](check-data.qmd).
