---
title: "Read your data"
description: "Turn a tracking export into an anipoint with aniread, and tell it about your recording."
order: 2
---

```{r}
#| label: setup
#| include: false
library(animovement)
knitr::opts_chunk$set(
  comment = "#>",
  fig.width = 7, fig.height = 4.5, fig.align = "center",
  dpi = 150, out.width = "100%"
)
```

[aniread](https://animovement.dev/aniread) reads movement data from a wide range
of sources: pose estimation from video, centroid trackers, motion capture, even
trackball hardware. Whichever tool wrote the file, you get back the same
structure, so after this first step it stops mattering.

## What it can read {#sec-sources}

These are the sources aniread supports, with the reader for each and the file
types it accepts:

```{r}
#| label: sources
get_supported_sources() |>
  print(n = Inf)
```

You rarely need to call those readers yourself. `read_dataset()` reads any of
these files: it works out which tool wrote it and calls the matching reader.

## Get some data to follow along {#sec-sample}

You don't need your own files. `get_sample_data()` downloads a real export for
any supported source and returns its path:

```{r}
#| label: sample
path <- get_sample_data("sleap")
basename(path)
```

We'll use this recording in all of the guides: ten minutes of a single mouse on
an elevated plus maze, tracked with [SLEAP](https://sleap.ai). To see what else
is available, call `get_sample_data(list_datasets = TRUE)`.

## Read it {#sec-read}

```{r}
#| label: read
movement <- read_dataset(path)
movement
```

## What you got back {#sec-anipoint}

That's an **anipoint**: one row per point (here, each keypoint of each
individual) at each moment. It's one of the *aniframe* classes defined by
[anicore](https://animovement.dev/anicore), which every other package in the
ecosystem reads and returns. Underneath it's a
[tibble](https://tibble.tidyverse.org), so `dplyr` and `ggplot2` work on it as
usual. On top, metadata records what each column is for:

| Question | Role | Read it with | Here |
|----------|------|--------------|------|
| **What** is moving? | identity | `get_variables(movement, "what")` | `individual`, `keypoint` |
| **When**? | index | `get_index(movement)` | `time` |
| **Where**? | position | `get_axes(movement)` | `x`, `y` |

```{r}
#| label: roles
get_variables(movement, "what")
get_index(movement)
get_axes(movement)
```

Two consequences worth knowing from the start:

- **Column names are free.** The roles are what matter, so a file whose identity
  column is called `track` works just as well.
- **The frame arrives grouped by identity**, so everything downstream runs
  within each track rather than across them:

```{r}
#| label: groups
dplyr::n_groups(movement)
```

Six groups: one mouse × six keypoints. That grouping is why interpolation
never bleeds from the snout into the tail, and why speed is never computed
across the join between two tracks.

::: {.callout-tip}
## Going deeper
This is the short version. anicore's articles cover the data structure in full:
[The aniframe data structure](https://animovement.dev/anicore/articles/aniframe-structure.html)
for the roles and how to set them, and
[Metadata on an aniframe](https://animovement.dev/anicore/articles/aniframe-metadata.html)
for everything below.
:::

## Tell it about your recording {#sec-metadata}

An anipoint keeps context that a bare data frame would lose. The reader fills in
what the file records:

```{r}
#| label: metadata
get_metadata(movement)
```

Two things are missing that everything downstream depends on: how fast the
camera recorded, and how big a pixel is.

### Time in seconds {#sec-time}

SLEAP's analysis files don't record a frame rate, so `time` counts frames.
Declare the rate, then convert the time column to seconds:

```{r}
#| label: set-rate
movement <- movement |>
  set_metadata(sampling_rate = 30) |>
  convert_unit_time("s")
```

::: {.callout-warning}
## Use your own frame rate
We're assuming 30 fps for this sample. Set whatever your recording actually
was: everything downstream that involves time scales with this number.
:::

### Space in centimetres {#sec-space}

Coordinates are in pixels (`unit_space: px`), which is all a video can tell
you. To get real distances you need something of known size in the picture.
Here we use the mouse itself: an adult mouse measures about 9 cm from its snout
to the base of its tail, and stretched out, this one spans about 140 pixels.

```{r}
#| label: calibrate
movement <- movement |>
  convert_unit_space("cm", calibration_factor = 9 / 140)
```

`convert_unit_space()` rescales every coordinate by that factor and records the
new unit. The header now shows the sampling rate and the duration, and from
here on, distances come out in centimetres and speeds in centimetres per
second:

```{r}
#| label: after-units
movement
```

::: {.callout-note}
## Calibrate against something in the same plane
For your own recordings, film an object of known size, such as a ruler or the
arena itself, at the height the animal moves at. A rough scale is fine for
following these guides, but measurements are only as good as the calibration
behind them.
:::

## Take a first look {#sec-look}

Calling `plot()` on an anipoint draws the trajectories, with axis labels taken
from the metadata:

```{r}
#| label: first-plot
#| fig-cap: "Raw SLEAP tracks. The plus maze is clearly visible, and so are a handful of points that landed nowhere near it."
plot(movement)
```

The maze shape comes straight out of the data, with no annotation needed. So do
the long lines reaching off to the left: a few dozen points, out of
`r format(nrow(movement), big.mark = ",")`, that landed well outside the maze.
They're drawn as long straight lines because the plot joins consecutive
positions. Whether that matters, and what else is hiding in there, is what the
[next guide](check-data.qmd) is for.

## The skeleton {#sec-structure}

Pose-estimation tools also record which keypoints are joined to which: the
skeleton. SLEAP keeps it in the same file as the tracks, so `read_dataset()`
has already attached it:

```{r}
#| label: structure
get_structure(movement)
```

With the skeleton attached, functions that work on body segments know what the
segments are. Some tools keep the skeleton in a separate file, such as
DeepLabCut's `config.yaml`. For those, read it with `read_structure()` and
attach it with `set_structure()`. The anicore article on
[structures](https://animovement.dev/anicore/articles/structures.html) covers
them in full.

## Reading your own files {#sec-own}

Point `read_dataset()` at your file. If you're curious what it decides the file
is, `detect_source()` tells you without reading it:

```{r}
#| label: detect
detect_source(path)
```

When detection can't tell two sources apart, or you want to be explicit, name
the source: `read_dataset(path, source = "sleap")`. Each reader's help page,
such as `?read_sleap`, documents the options specific to that format. Pass them
to `read_dataset()` and it hands them on.

For a format aniread doesn't cover, `read_custom()` maps arbitrary columns onto
the roles above. And once you've read something, `write_aniframe()` saves it to
a Parquet file, metadata and all, which `read_aniframe()` opens again far
faster than re-parsing the original export.

Next: [check what you actually have](check-data.qmd).
